Motion planning for assessing accessibility of protein tunnels

### 2017-09-10 paper.tex
method

In this paper, we propose a modification to RRT to address the problem of planning a non-spherical flexible ligand passage through a given protein.
We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
The potential tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
This optimization allows us to find trajectories along the given tunnel, which is faster than searching for any trajectory in the whole protein.
To enable the movement of the ligand in the narrow parts of the tunnel, the free-space can be dilated by shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}. 
The ligand flexibility is modeled using a fixed set of predefined configurations which the ligand can occupy.

@@

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.

@@

\section{Proposed Method}
To address the problems of the existing solutions for computing the non-spherical flexible ligand trajectories through protein tunnels, we propose a modification of the RRT planner. 
Our approach takes into account the shape and flexibility of the ligand and localizes the searched configuration space using the information about tunnels detected by the Voronoi-based approach.
The input is therefore formed by a set of tunnels computed using Voronoi diagrams and the task is to find trajectories for a ligand along these tunnels
leading from outside the protein to the active site.
The tunnels are detected in a sequence time steps capturing protein dynamic behavior that is computed using molecular dynamics simulations.

@@

Due to the protein dynamics, the tunnel changes, move and merge, and their bottleneck changes in each frame.
It can therefore be expected, that the detected tunnels are rather narrow.
Based on the observation from MD simulations, the tunnels adapt to the ligands (and vice versa), so even narrow tunnels can serve as transportation path for thee ligands TODO CITACE??.
To enable computing of trajectories in the narrow tunnels, we shrink the atom radii of the ligand, similarly to~\cite{cortes2010simulating,guieysse2008structure}.

The ligand flexibility is modeled using a predefined set of typical conformations, therefore the number of the dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
While classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
This metric measures distances between two configurations as the nearest distance of two atoms. 
This enables the metric to consider the shape of the ligand and it supports retraction of the ligand to the narrow passages.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
The tunnels can be found by the tunnel detection tools like CAVER 3.0~\cite{caver3}.

@@

%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The flexibility of a ligand is modeled using the set $\L$ of its conformations.
We assume that a library of all possible (in the considered resolution) conformations is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for motion planning.
%The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.

To enable motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
A discrete set of scales is used, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (the geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$ and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation  parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

\subsection{RRT for computing ligands trajectory along a tunnel}

The proposed motion planning is based on the principle of the RRT~\cite{lavalleRRT} planner.
In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric computed using
both 3D rotation and 3D translation.
The main loop of the planner (Alg.~\ref{alg::main}) is similar to original RRT, but we propose three modifications.

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
This allows the tree to slightly detour from the tunnel and overcome its difficult (narrow) parts by finding an alternative trajectory nearby.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
metric for the nearest-neighbor search.
The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
of $\rv$ cause  exploration of paths around the tunnel.
We propose to set this parameter to the average width of the tunnel.

\linesnumbered
\begin{algorithm}[h]
{\small
\setstretch{0.88}
\caption{\label{alg::main}Main loop of the RRT planner}
\KwIn{
    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
    initial configuration $\qinit$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$,
   distance $\dt$ to move virtual goal
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
$\T$.addNode($\qinit$)\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\T, \qnear,\qrand$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $c_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
}
\end{algorithm}

The second proposed modification is the expansion procedure (Alg.~\ref{alg::expand}), which expands the tree from $\qnear$ towards $\qrand$.
The expansion towards $\qrand$ is crucial for the growth of the tree towards unexplored areas of the configuration space.
Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
distance $\varepsilon$ from $\qnear$.
This would however mostly lead to a collision in the case of ligands.

The expansion procedure attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure preffers expanding the tree by larger scales.
This forces the planner to automatically use smaller scaled in the difficult (narrow) parts of the tunnels, while prefering
large scales if possible.

The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
This metric is denoted $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
$\da(q_1,q_2)$.
By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
is useful when multiple conformations are utilized.
By employing this metric in the expansion procedure, the tree is expanded by such a new configuration, which atoms maximally approach
the atoms of the configuration $\qrand$.
We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability
of reaching the active site.

@@

%Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration $q$ being tested.
The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
is also the nearest point of the ligand at $q$ to $\qrand$.

\begin{algorithm}[h]
{\small
\setstretch{0.88}
\caption{\label{alg::expand}expand}
\KwIn{
   tree $\T$,
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$, and $\sdelta$,
}
%\KwOut{
%    set $S$ of collision-free configurations reachable from $\qnear$\;
%}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
        $\qnew = \emptyset$; // empty configuration \nllabel{alg::expa} \\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \nllabel{alg::expb}
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to the next conformation
        }
    }
}
}
\end{algorithm}

@@

In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are close to each other, but 
the atoms can also form a more longer shape (Fig.~\ref{fig::m040c}).

From motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in a narrow passage, while the packed conformations better rotate at a place.
In the narrow parts of the tunnels, the tree can be extended by the longer conformations, while in the more open regions, also the
packed configurations can be used.

@@

\begin{figure}
\centering
%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

The proposed approach was used to analyze tunnel traversability towards the active site of Haloalkane dehalogenase protein (PDB ID 1CQW).
The active site was defined using residues 38 (Asparagine), 102 (Histidine) and 103 (Aspartate).
The molecular dynamics of the protein was computed using the Amber tool.
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.
The tunnel traversability was analyzed in 100 frames.

>>>>

In this paper, we propose a modification to RRT to address the problem of planning a non-spherical flexible ligand passage through a given protein.
We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
The potential tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
This optimization allows us to find trajectories along the given tunnel, which is faster than searching for any trajectory in the whole protein.
To enable the movement of the ligand in the narrow parts of the tunnel, the free-space can be dilated by shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}. 
The ligand flexibility is modeled using a fixed set of predefined conformations which the ligand can occupy.

@@

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.

@@

\section{Proposed Method}
To address the problems of the existing solutions for computing the non-spherical flexible ligand trajectories through protein tunnels, we propose a modification of the RRT planner. 
Our approach takes into account the shape and flexibility of the ligand and localizes the searched configuration space using the information about tunnels detected by the Voronoi-based approach.
The input is therefore formed by a set of tunnels computed using Voronoi diagrams and the task is to find trajectories for a ligand along these tunnels
leading from outside the protein to the active site.
The tunnels are detected in a sequence time steps capturing protein dynamic behavior that is computed using molecular dynamics (MD) simulations.

@@

Due to the protein dynamics, the tunnel changes, move and merge, and their bottleneck can change in each time step.
Therefore, it can be expected that the detected tunnels are rather narrow.
Based on the observations of MD simulations, tunnels adapt their shape to the passing ligands and vice versa.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand approximating sphere, can serve as the transportation path for this ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates shrinking of the ligand size.
In our solution we simulate this by shrinking the atom radii of the ligand, used also by~\cite{cortes2010simulating,guieysse2008structure}.

Molecular dynamics simulations of ligands passing through proteins show that the ligands can be highly flexible.
However, ligands cannot form arbitrary conformations. 
Ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.

While the classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
This metric measures the distances between two configurations as the nearest distance of two atoms. 
This enables the metric to consider the shape of the ligand and it supports the retraction of the ligand to the narrow passages.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for our proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.

@@

%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The flexibility of the ligand is modeled using the set $\L$ of its conformations.
We assume that the library of all possible conformations in a considered resolution is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for the motion planning.
%The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.

To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$, and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

\subsection{RRT for Computing Ligand's Trajectory along Tunnel}

The proposed motion planning is based on the principle of the RRT~\cite{lavalleRRT} planner.
In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric computed using
both 3D rotation and 3D translation.
The main loop of the planner (Alg.~\ref{alg::main}) is similar to the original RRT, but we propose its three modifications.

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
This allows the tree to slightly detour from the tunnel and overcome its difficult (i.e., narrow) parts by finding an alternative trajectory nearby.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

\linesnumbered
\begin{algorithm}[bh]
{\small
\setstretch{0.88}
\caption{\label{alg::main}Main loop of the RRT planner}
\KwIn{
    tunnel $T=( (c_i, r_i) )$, $i=1,\ldots,n$, with spheres centers $c_i \in \R^3$ and radii $r_i$,
    initial configuration $\qinit$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 1$; // index of the virtual goal\\
$iteration = 0$\;
$\T$.addNode($\qinit$)\;
\While{$iteration < \Imax$ {\bf and}  $v < n$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around virtual goal $c_v\!\in\!T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;
    expand($\qnear,\qrand$)\;
    \For{$i= n-1,n-2,\ldots,v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree towards sphere $c_i$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
}
\end{algorithm}

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be set to zero, as these are not used in the employed
metric for the nearest-neighbor search.
The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
of $\rv$ cause the exploration of paths around the tunnel.
We propose to set this parameter to the average width of the tunnel.

The second proposed modification is the expansion procedure (Alg.~\ref{alg::expand}), which expands the tree from $\qnear$ towards $\qrand$.
The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
distance $\varepsilon$ from $\qnear$.
However, in the case of ligands, this would mostly lead to a collision.

The expansion procedure attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales.
This forces the planner to automatically use smaller scales in the difficult (narrow) parts of the tunnels, while preferring
large scales if possible.

The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
$\da(q_1,q_2)$.
By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
is useful when multiple conformations are utilized.
By employing this metric in the expansion procedure, the tree is expanded by such a new configuration, whose atoms maximally approach
the atoms of the configuration $\qrand$.
We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability
of reaching the active site.

@@

%Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration $q$ being tested.
The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$, and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
is also the nearest point of the ligand at $q$ to $\qrand$.

\begin{algorithm}[th]
{\small
\setstretch{0.88}
\caption{\label{alg::expand}expand}
\KwIn{
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$, and $\sdelta$,
   tree $\T$
}
%\KwOut{
%    set $S$ of collision-free configurations reachable from $\qnear$\;
%}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
        $\qnew = \emptyset$; // empty configuration \nllabel{alg::expa} \\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \nllabel{alg::expb}
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to the next conformation
        }
    }
}
}
\end{algorithm}

@@

In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).

From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
packed conformations can be used.

\begin{figure}[b!]
\centering
%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

@@

The proposed approach was used to analyze tunnel traversability towards the active site of the Haloalkane dehalogenase protein (PDB ID 1CQW).
The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
The molecular dynamics of the protein was computed using the AMBER tool~\cite{amber}.
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck of the tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the first and the second tunnel, resp.
The tunnel traversability was analyzed in 100 frames.
